---
title: "1_data_wrangling"
author: "..."
date: "2025-10-01"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)

base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',  #log files go here
                      message=FALSE, warning=FALSE,
                      cache = TRUE) 
```  


# Analysis Scripts

- **Description:**
    - Create two analysis scripts:
        - **Script 1:** Initial descriptions of the breast cancer data set
        - **Script 2:** Implement random forest algorithm using tidy models and R
- **Code Organization:**
    - Separate code for each script and analysis


```{r}
# Load libraries
library(dplyr)
```

## Load Data
```{r}
#load data
raw_dat <- read.csv("../_data/Breast Cancer Wisconsin Data.csv")
str(raw_dat)
```

## Descriptive Summary of Data
```{r}
mutate_dat <- 
  raw_dat %>% mutate(diag_text = ifelse(diagnosis=="B", "Benign", "Malignant")) %>%  
  group_by(diag_text) %>%
  summarise(
    count = n(),
    mean_radius = mean(radius_mean, na.rm = TRUE),
    mean_texture = mean(texture_mean, na.rm = TRUE)
  )
```
## Viz
```{r}
library(ggplot2)
#regression view of the diagnosis types

raw_dat %>% mutate(diag_text = ifelse(diagnosis=="B", "Benign", "Malignant")) %>%  
  ggplot(aes(x=radius_mean, y =  texture_mean, size = area_mean, color=diag_text)) +
  geom_point(size=3) +
  geom_smooth(method="lm")+
  labs(title="Mean Radius vs Mean Texture by Diagnosis Type",
       x="Mean Radius",
       y="Mean Texture",
       color="Diagnosis Type") +
  theme_minimal()

```
### Comment-CTA:: 
- Already we are seeing reasonable separation between the BENING and MALIGNANT groups based on just a few of these variables...
- Dataset likely would need dimens reduction.


## Minimal Random Forest Implementation
```{r}
library(tidymodels)
#install.packages("ranger")
# Set seed for reproducibility
set.seed(123)
```

```{r}
# Split the data into training and testing sets
data_split <- initial_split(raw_dat, prop = 0.8, strata = diagnosis)
train_data <- training(data_split)
test_data <- testing(data_split)
```

```{r}
# Define the random forest model
rf_model <- rand_forest(mtry = 5, trees = 1000, min_n = 10) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

```{r}
# Create a recipe for preprocessing
rf_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  step_rm(id) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# Create a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)
```

```{r}
# Train the model
rf_fit <- rf_workflow %>%
  fit(data = train_data)
```

```{r}
# Make predictions on the test set
rf_predictions <- rf_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(diagnosis))
```

```{r}
# Evaluate the model
rf_metrics <- rf_predictions %>% mutate(diagnosis = factor(diagnosis, levels = c("B", "M"))) %>%
  metrics(truth = diagnosis, estimate = .pred_class)

print(rf_metrics)
```

## Plot to show Prediction differential
```{r}
rf_predictions %>%
  ggplot(aes(x = .pred_class, fill = diagnosis)) +
  geom_histogram(stat = "count", position = "identity", alpha = 0.6, bins = 30) +
  labs(title = "Random Forest Predictions: Probability of Malignant",
       x = "Predicted Probability of Malignant",
       y = "Count",
       fill = "Actual Diagnosis") +
  theme_minimal()
```

