---
title: "01_random_forest_model"
author: "You!"
date: "2025-10-01"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE) #sets all knit options!! base_dir, base_url, fig_path
```  


# Analysis Scripts::

- **Description:**
    - Create two analysis scripts:
        - **Script 1:** Initial descriptions of the breast cancer data set
        - **Script 2:** Implement random forest algorithm using tidy models and R
- **Code Organization:**
    - Separate code for each script and analysis


```{r libraries}
# Load libraries
library(dplyr)
library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(ggplot2)   # For visualization
```

### Load & Prep Data
```{r}
#load data
raw_BCWDD <- read_csv(here("_data", "Breast Cancer Wisconsin Data.csv"))
names(raw_BCWDD) <- gsub(names(raw_BCWDD), pattern = " ", replacement = "_")
str(raw_BCWDD)
```

## Descriptive Summary of Data
```{r}
cancer_data <- raw_BCWDD %>% 
  mutate(
    # Convert the diagnosis column to a factor
    # This is ESSENTIAL for classification models in tidymodels
    diagnosis = as_factor(diagnosis)) %>% 
    #diagnosis.fac_lab = factor(diagnosis, levels = c("B", "M"), labels=c("B", "M"))) # 'B' = Benign, 'M' = Malignant
  select(-33) #remove unnamed column if present
```

## Viz
```{r}
#regression view of the diagnosis types
cancer_data %>% mutate(diag_text = ifelse(diagnosis == "B", "Benign", "Malignant")) %>%
  ggplot(aes(x=radius_mean, y =  texture_mean, color=diag_text)) +
  geom_point(aes(size=area_mean)) +
  geom_smooth(method="lm")+
  labs(title="Mean Radius vs Mean Texture by Diagnosis Type",
       x="Mean Radius",
       y="Mean Texture",
       color="Diagnosis Type") +
  theme_minimal()

```

Visualization of three variables; Radius, Texture, Size - and there is already a reasonable visual separation between Benign and Malignant diagnosis. However, there is significant cross over in the middle of the chart. A typical regression model would have difficulty classifying cases in this area - and similarly a person trying to decide if they should operate on a potential cancerous growth would be in a coinflip situation. 

This is where machine learning (particularly non-linear methods) that can incorporate more variables will start to shine in distinguishing the close match cases. However, visual intuition is often a good marker for how much we can undertand about how a model will make decisions. (Rule of 3). If we added in shapes or opacity based on other variables, the chart would become confusing, even as it might slightly emphasize the distinctino between Bening and Malignant cases. 

```{r}
#same chart - RANGE restricted to radius_mean 10:20 and texture_mean 15:30.
cancer_data %>% mutate(diag_text = ifelse(diagnosis == "B", "Benign", "Malignant")) %>%
  dplyr::filter(radius_mean >=10 & radius_mean <=20 & texture_mean >=15 & texture_mean <=30) %>%
  ggplot(aes(x=radius_mean, y =  texture_mean, color=diag_text)) +
  geom_point(aes(size=area_mean, alpha=compactness_mean)) +
  geom_smooth(method="lm")+
  labs(title="Mean Radius vs Mean Texture by Diagnosis Type (Zoomed In)",
       x="Mean Radius",
       y="Mean Texture",
       color="Diagnosis Type") +
  theme_minimal()
```
The updated chart is focused on the 'decision boundary' that will be explored in the ML models. Only one more variable has been added - compactness - mapped to opacity. The decision boundary is still quite fuzzy at best!



## Minimal Random Forest Implementation
```{r}
# Set seed for reproducibility
set.seed(8888)
```

```{r data-split and k-folds}
# Split the data into training and testing sets
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)
# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)
```

## fit a single model
```{r rf-model-engine}
# Define the random forest model
rf_model <- rand_forest(mtry = 5, trees = 1000, min_n = 10) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

```{r recipe-preprocess}
# Create a recipe for preprocessing
rf_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  step_rm(id) %>%
  step_normalize(all_numeric_predictors())
```

```{r workflow-setup}
# Create a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)
```

```{r training-fitting}
# Train the model
rf_fit <- rf_workflow %>%
  fit(data = train_data)
```

```{r model-predictions}
# Make predictions on the test set
rf_predictions <- rf_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(diagnosis))
```

```{r model-metrics}
# Evaluate the model
rf_metrics <- rf_predictions %>% mutate(diagnosis = factor(diagnosis, levels = c("B", "M"))) %>%
  metrics(truth = diagnosis, estimate = .pred_class)

print(rf_metrics)
```

## Plot to show Prediction differential
```{r}
rf_predictions %>%
  ggplot(aes(x = .pred_class, fill = diagnosis)) +
  geom_histogram(stat = "count", position = "identity", alpha = 0.6, bins = 30) +
  labs(title = "Random Forest Predictions: Probability of Malignant",
       x = "Predicted Probability of Malignant",
       y = "Count",
       fill = "Actual Diagnosis") +
  theme_minimal()
```
**Comments** 
- This bar plot is showing the model predictions vs accuracy. Essentially, the SMALL OVERLAP in the light Blue (M) for malignant is showing where the model classified a BENIGN tumor as Malignant. 


## Machine Learning Routine

- This is the same as above, but a **workflow** is put into place to evaluate the (already quite good) random forest algorithm across multiple parameters. This is the best practice for improving accuracy. 
```{r}
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)

```

