---
title: "01_random_forest_model"
date: "2025-10-01"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE) #sets all knit options!! base_dir, base_url, fig_path
```  


## Random Forest Implementation
- Implement RF Model
- Use cross Validation
- Save model outpu

```{r libraries}
# Load libraries
library(dplyr)
library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(ggplot2)     # For visualization
library(vip)         #variables of importance
set.seed(8888) 
```

### Load & Prep Data
```{r}
#load data
cancer_data <- readRDS(here("_data", "cancer_data_processed.rds")) #processed in00 data_wrangling script.
str(cancer_data)
```

### 2. Data Splitting
```{r data-split and k-folds}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

## fit a single model
```{r rf-model-engine}
# Define the random forest model
rf_model <- rand_forest(mtry = 5, trees = 1000, min_n = 10) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

```{r recipe-preprocess}
# Create a recipe for preprocessing
rf_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  step_rm(id) %>%
  step_normalize(all_numeric_predictors())
```

```{r workflow-setup}
# Create a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)
```

```{r training-fitting}
# Train the model
rf_fit <- rf_workflow %>%
  fit(data = train_data)
```

```{r model-predictions}
# Make predictions on the test set
rf_predictions <- rf_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(diagnosis))
```

```{r model-metrics}
# Evaluate the model
rf_metrics <- rf_predictions %>% mutate(diagnosis = factor(diagnosis, levels = c("B", "M"))) %>%
  metrics(truth = diagnosis, estimate = .pred_class)

print(rf_metrics)
```

## Plot to show Prediction differential
```{r}
rf_predictions %>%
  ggplot(aes(x = .pred_class, fill = diagnosis)) +
  geom_histogram(stat = "count", position = "identity", alpha = 0.6, bins = 30) +
  labs(title = "Random Forest Predictions: Probability of Malignant",
       x = "Predicted Probability of Malignant",
       y = "Count",
       fill = "Actual Diagnosis") +
  theme_minimal()
```
**Comments** 
- This bar plot is showing the model predictions vs accuracy. Essentially, the SMALL OVERLAP in the light Blue (M) for malignant is showing where the model classified a BENIGN tumor as Malignant. 


## Machine Learning Routine

- This is the same as above, but a **workflow** is put into place to evaluate the (already quite good) random forest algorithm across multiple parameters. This is the best practice for improving accuracy. 
```{r}
#set.seed(8888)
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### Model Specification for Tuning

We will tune `mtry` (number of predictors per split) and `min_n` (minimum node size). We'll set the engine to `ranger` for speed and to enable variable importance.

```{r rf-spec-tune}
# Define the random forest model, but with tunable parameters
rf_spec_tune <- rand_forest(
  mtry = tune(),
  trees = 1000, # 1000 trees is usually enough, no need to tune
  min_n = tune()
) %>%
  set_engine("ranger", importance = "permutation") %>% # Set importance for vip
  set_mode("classification")
```

### Recipe and Workflow

Update the recipe with the tuning step, and tunable model (ranger model)

```{r rf-workflow-tune}
rf_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it
  step_normalize(all_numeric_predictors())

# Create a new workflow for *tuning*
rf_workflow_tune <- workflow() %>%
  add_model(rf_spec_tune) %>%
  add_recipe(rf_recipe)
```

### Hyperparameter Tuning

Now we run the tuning process across our 10 CV folds.

```{r rf-run-tuning}
# Create a grid of hyperparameters to try
rf_grid <- grid_latin_hypercube(
  mtry(range = c(1, 30)), # Try values for mtry from 1 to 30
  min_n(range = c(2, 20)), # Try values for min_n from 2 to 20
  size = 20
)

# Run the tuning
rf_tune_results <- tune_grid(
  rf_workflow_tune,
  resamples = cv_folds,
  grid = rf_grid,
  #metrics = metric_set(accuracy, precision, recall),
  control = control_grid(save_workflow=TRUE)
)

# Show the top 5 best-performing hyperparameter sets
show_best(rf_tune_results, metric = "accuracy")

# Select the single best set of parameters
best_rf_params <- select_best(rf_tune_results, metric = "accuracy")

#
```

### Finalize and Evaluate Model

With our best parameters, we do a `last_fit()` to get our final, unbiased test metrics.

```{r rf-final-fit}
# Finalize the workflow with the best parameters
final_rf_workflow <- finalize_workflow(
  rf_workflow_tune,
  best_rf_params
)

# Use 'last_fit' to fit on the full training set and evaluate on the test set
final_rf_fit <- last_fit(
  final_rf_workflow,
  data_split
)
```

### Review Performance Metrics

Let's see how our *tuned* model performed on the test data.

```{r rf-final-metrics}
# Get the metrics from our test-set evaluation
test_metrics <- collect_metrics(final_rf_fit)

# Get the predictions to build a confusion matrix
test_predictions <- collect_predictions(final_rf_fit)

# Generate and print the confusion matrix
conf_matrix <- conf_mat(
  test_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") # Set positive = "M"
)

print(conf_matrix)
autoplot(conf_matrix, type = "heatmap")
```

### Feature Importance

This is a key benefit of Random Forest. We can see which variables the model used most to make its decisions.

```{r rf-vip}
#library(vip)
# Extract the fitted model object
fitted_rf_model <- extract_workflow(final_rf_fit)
best_rf_fit <- fit_best(rf_tune_results)

# Use the 'vip' package to create a variable importance plot
print("Plotting Variable Importance...")
vip(fitted_rf_model, num_features = 20) +
  labs(title = "Random Forest Variable Importance")
```

### Save Model Artifacts
```{r rf-save-artifacts}
# Save the final, fitted workflow
saveRDS(
  list("model" = final_rf_fit, "metrics" = test_metrics, "preds" = test_predictions),
  file = here("_models", "final_rf_output.rds")
)

## Save the test metrics
#saveRDS(
#  test_metrics,
#  file = here("_models", "rf_test_metrics.rds")
#)
```
