---
title: "02_lasso_regression.md"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)
```  

## OUTLINE OF ANALYSIS 

Proposed Method 2: 
  - (Implement Lasso): Implement Lasso Logistic Regression. 
  - Use Lasso (L1) Logistic Regression, specifically to see if a sparse model can be interpretable for this dataset without sacrificing too much accuracy. 
  - Use cross-validation to select the optimal regularization parameter (lambda). 
  - The resulting model's coefficients will be analyzed to produce a ranked list of the most important diagnostic features.
  
## Lasso Regression Implementation
### Methodological Note: 
Lasso (Least Absolute Shrinkage and Selection Operator) extends standard Logistic Regression by adding an $L1$ penalty to the objective function. For a binary outcome $y_i \in \{0,1\}$ (Benign/Malignant), we minimize the negative log-likelihood with a constraint on the sum of the absolute values of the coefficients:
$$ \min_{(\beta_0, \beta)} \left{ - \frac{1}{N} \sum_{i=1}^{N} [y_i (x_i^T \beta + \beta_0) - \log(1 + e^{x_i^T \beta + \beta_0})] + \lambda \sum_{j=1}^{p} |\beta_j| \right} $$

Where:$\lambda$ is the tuning parameter that controls the strength of the penalty.As $\lambda$ increases, coefficients $\beta_j$ are shrunk towards zero.Feature Selection: Unlike Ridge regression, Lasso can shrink coefficients to exactly zero, effectively removing those variables from the model. This makes Lasso ideal for this dataset, as it identifies the minimal subset of tumor features required for accurate diagnosis, enhancing interpretability.


```{r libraries}

# 1. SETUP LIBRARIES
library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(glmnet)      # The 'engine' for penalized regression
library(vip)         # For variable importance plots
library(conflicted)  # To manage function name conflicts

# Set seed: 
set.seed(8888) 
```

### 1. Load and Prepare Data

```{r}
#load data
cancer_data <- readRDS(here("_data", "cancer_data_processed.rds")) #processed in00 data_wrangling script.
str(cancer_data)
```
### 2. Data Splitting

We use the same stratified split parameters and seed as the SVM script to ensure the train/test sets are identical.

```{r data-split}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

### 3. Preprocessing Recipe

This recipe is nearly identical to the SVM one. Scaling and centering are **critical** for Lasso regression, as the penalty is applied directly to the coefficients' values.

Critical Step: Lasso requires all predictors to be on the same scale because the penalty term $\sum |\beta_j|$ treats all coefficients equally. If one variable has a range of 0-1000 and another 0-1, the larger one would be penalized unfairly without normalization.

```{r recipe}
# 4. PREPROCESSING RECIPE
# We are predicting 'diagnosis' using all other variables (.).
# The only predictor we *don't* want is 'id', so we remove it.
# We then normalize (center and scale) ALL numeric predictors.
lasso_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it
  step_normalize(all_numeric_predictors()) # Center and scale

```

### 4. Cross-Validation Folds

We create the *exact same* 10-fold cross-validation folds from our training data, as the `set.seed()` was identical.

```{r cv-folds}
# 5. CROSS-VALIDATION FOLDS
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### 5. Model Specification & Tuning

Here we define the `logistic_reg` model. We set `mixture = 1` to specify *pure Lasso* (L1 penalty) and tell `tidymodels` we want to `tune()` the `penalty` (lambda) parameter.

- We tune penalty (lambda). We set mixture = 1 to define a pure Lasso model (mixture=0 is Ridge, mixture=0.5 is Elastic Net).

```{r model-spec}
# 6. MODEL SPECIFICATION
# We specify a logistic regression model.
# 'mixture = 1' makes it a pure Lasso model (L1 penalty).
# We set the 'engine' to 'glmnet'.
# We will tune the 'penalty' (lambda) hyperparameter.
lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# 7. CREATE WORKFLOW
# Combine the recipe and model spec
lasso_workflow <- workflow() %>%
  add_recipe(lasso_recipe) %>%
  add_model(lasso_spec)
```


```{r}
# 8. HYPERPARAMETER TUNING
# Create a grid of 50 different 'penalty' (lambda) values to try
lasso_grid <- grid_regular(penalty(), levels = 50)
#ctrl <- control_grid(save_workflow = TRUE)
# Run the tuning process
lasso_tune_results <- tune_grid(
  lasso_workflow,
  resamples = cv_folds,
  grid = lasso_grid,
  control = control_grid(save_workflow = TRUE),
  metrics = metric_set(accuracy, precision, recall))

# Show the top 5 best-performing penalty values
show_best(lasso_tune_results, metric = "accuracy")

# Select the single best penalty value
best_lasso_params <- select_best(lasso_tune_results, metric = "accuracy")
best_lasso_params
```

```{r}
#fit best model ^^ this is duplication of code from above, but worth seeing/saving the params. Can also fit with the params
best_lasso_fit <- fit_best(lasso_tune_results)
tidy(best_lasso_fit)
```

### 6. Finalize and Evaluate Model

We update our workflow with the *best* penalty and fit it to the full training set, then evaluate on the test set.

```{r final-fit}
# 9. FINALIZE WORKFLOW
# Update the workflow with the best penalty we just found
# same as best_lasso_fit...
final_lasso_workflow <- finalize_workflow(lasso_workflow,
  #lasso_workflow,
  best_lasso_params
) 

# 10. FIT ON TRAIN, EVALUATE ON TEST
# Use 'last_fit' to fit on the full training set and evaluate on the test set
final_lasso_fit <- last_fit(
  final_lasso_workflow,
  data_split
)
```

### 7. Review Performance Metrics

Let's see how our final tuned Lasso model performed on the test data.
## Model Metrics
```{r metrics}
# 11. COLLECT METRICS
# Get the metrics from our test-set evaluation
test_metrics <- collect_metrics(final_lasso_fit)
print(test_metrics)
```

## Accuracy, Sensitivity, Recall Metrics

```{r}
# Get the predictions to build a confusion matrix
test_predictions <- collect_predictions(final_lasso_fit)
ml_mets <- metric_set(accuracy, precision, recall, f_meas)
ml_mets(test_predictions, truth = diagnosis, estimate = .pred_class)
```

## Generate and print the confusion matrix
```{r}
# We again set positive = "M" for consistency
conf_matrix <- conf_mat(
  test_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") 
)

print(conf_matrix)
```

```{r}
autoplot(conf_matrix, type = "heatmap")
```

```{r}
# Precision-Recall Curve
# Compare this to the Random Forest curve to see which model
# maintains higher precision as recall increases.
test_predictions %>%
  roc_curve(diagnosis, .pred_M) %>%
  autoplot() +
  labs(
    title = "ROC Curve: Lasso Regression",
    subtitle = "Performance of the L1-regularized Logistic Model"
  )
```

### 8. Feature Importance (Interpretability)

This is the key step for the Lasso model. We will extract the coefficients from our final model to see which features it "kept" (gave a non-zero coefficient) and which it "threw away" (set to zero).

```{r vip}
# 12. EXTRACT FEATURE IMPORTANCE
#library(vip)
# Extract the fitted model object
fitted_lasso_model <- extract_workflow(final_lasso_fit)

# Use the 'vip' package to create a variable importance plot
# This shows the non-zero coefficients
print("Plotting Variable Importance...")
vip(fitted_lasso_model, num_features = 20) +
  labs(title = "Lasso Variable Importance",
       subtitle = "Features with non-zero coefficients")
```

```{r}
# We can also 'tidy' the model to see the coefficients as a data frame
print("Model Coefficients:")
tidy(best_lasso_fit) %>%
  dplyr::filter(estimate != 0) %>%  # Filter to show only non-zero features
  arrange(desc(abs(estimate))) # Sort by importance
```

### 9. Save Model Artifacts
```{r save-artifacts}
# 13. SAVE ARTIFACTS
# Save the final, fitted workflow
saveRDS(
  list("model" = final_lasso_fit, "metrics" = test_metrics, "preds" = test_predictions),
  file = here("_models", "final_lasso_output.rds")
)

```
