---
title: "02_lasso_regression.md"
author: "You!"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)
# 
# base_dir <- "../" # where output should default to
# base_url <- "" # keep as is
# fig_path <- "_plot_images/" # where image output should go
# 
# knitr::opts_knit$set(base.dir = base_dir)
# knitr::opts_chunk$set(fig.path = fig_path,
#                       cache.path = '../cache/',  #log files go here
#                       message=FALSE, warning=FALSE,
#                       cache = TRUE) 
```  

## OUTLINE OF ANALYSIS 

Proposed Method 2: 
  - (Implement Lasso): Implement Lasso Logistic Regression. 
  - Use Lasso (L1) Logistic Regression, specifically to see if a sparse model can be interpretable for this dataset without sacrificing too much accuracy. 
  - Use cross-validation to select the optimal regularization parameter (lambda). 
  - The resulting model's coefficients will be analyzed to produce a ranked list of the most important diagnostic features.


```{r setup, include=FALSE}

# 1. SETUP LIBRARIES
library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(glmnet)      # The 'engine' for penalized regression
library(vip)         # For variable importance plots
library(conflicted)  # To manage function name conflicts

# Set seed: 
set.seed(8888) 
```

### 1. Load and Prepare Data

Load the cleaned data and convert `diagnosis` to a factor for classification.

```{r load-data}
# 2. LOAD DATA
# Load the cleaned data from the _data folder
raw_BCWDD <- read_csv(here("_data", "Breast Cancer Wisconsin Data.csv"))

names(raw_BCWDD) <- gsub(names(raw_BCWDD), pattern = " ", replacement = "_")

cancer_data <- raw_BCWDD %>% 
  mutate(
    # Convert the diagnosis column to a factor
    # This is ESSENTIAL for classification models in tidymodels
    diagnosis = as_factor(diagnosis)) %>% 
    #diagnosis.fac_lab = factor(diagnosis, levels = c("B", "M"), labels=c("B", "M"))) # 'B' = Benign, 'M' = Malignant
  select(-33) #remove unnamed column if present


#str(cancer_data$diagnosis)
#str(cancer_data$diagnosis.fac_lab)

glimpse(cancer_data)
```

### 2. Data Splitting

We use the same stratified split parameters and seed as the SVM script to ensure the train/test sets are identical.

```{r data-split}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

### 3. Preprocessing Recipe

This recipe is nearly identical to the SVM one. Scaling and centering are **critical** for Lasso regression, as the penalty is applied directly to the coefficients' values.

```{r recipe}
# 4. PREPROCESSING RECIPE
# We are predicting 'diagnosis' using all other variables (.).
# The only predictor we *don't* want is 'id', so we remove it.
# We then normalize (center and scale) ALL numeric predictors.
lasso_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it
  step_normalize(all_numeric_predictors()) # Center and scale

```

### 4. Cross-Validation Folds

We create the *exact same* 10-fold cross-validation folds from our training data, as the `set.seed()` was identical.

```{r cv-folds}
# 5. CROSS-VALIDATION FOLDS
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### 5. Model Specification & Tuning

Here we define the `logistic_reg` model. We set `mixture = 1` to specify *pure Lasso* (L1 penalty) and tell `tidymodels` we want to `tune()` the `penalty` (lambda) parameter.

```{r model-spec}
# 6. MODEL SPECIFICATION
# We specify a logistic regression model.
# 'mixture = 1' makes it a pure Lasso model (L1 penalty).
# We set the 'engine' to 'glmnet'.
# We will tune the 'penalty' (lambda) hyperparameter.
lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# 7. CREATE WORKFLOW
# Combine the recipe and model spec
lasso_workflow <- workflow() %>%
  add_recipe(lasso_recipe) %>%
  add_model(lasso_spec)

# 8. HYPERPARAMETER TUNING
# Create a grid of 50 different 'penalty' (lambda) values to try
lasso_grid <- grid_regular(penalty(), levels = 50)
ctrl <- control_grid(save_workflow = TRUE)
# Run the tuning process
lasso_tune_results <- tune_grid(
  lasso_workflow,
  resamples = cv_folds,
  grid = lasso_grid,
  control = ctrl,
  metrics = metric_set(accuracy, precision, recall))

# Show the top 5 best-performing penalty values
show_best(lasso_tune_results, metric = "accuracy")

# Select the single best penalty value
best_lasso_params <- select_best(lasso_tune_results, metric = "accuracy")

#fit best model ^^ this is duplication of code from above, but worth seeing/saving the params. Can also fit with the params
best_lasso_fit <- fit_best(lasso_tune_results)
```

### 6. Finalize and Evaluate Model

We update our workflow with the *best* penalty and fit it to the full training set, then evaluate on the test set.

```{r final-fit}
# 9. FINALIZE WORKFLOW
# Update the workflow with the best penalty we just found
# same as best_lasso_fit...
final_lasso_workflow <- finalize_workflow(lasso_tune_results,
  #lasso_workflow,
  best_lasso_params
) 

wflow_fit <- fit(final_lasso_workflow, train_data)

# 10. FIT ON TRAIN, EVALUATE ON TEST
# Use 'last_fit' to fit on the full training set and evaluate on the test set
final_lasso_fit <- last_fit(
  final_lasso_workflow,
  data_split
)
```

### 7. Review Performance Metrics

Let's see how our final tuned Lasso model performed on the test data.

```{r metrics}
# 11. COLLECT METRICS
# Get the metrics from our test-set evaluation
test_metrics <- collect_metrics(final_lasso_fit)
print(test_metrics)

# Get the predictions to build a confusion matrix
test_predictions <- collect_predictions(final_lasso_fit)

# Generate and print the confusion matrix
# We again set positive = "M" for consistency
conf_matrix <- conf_mat(
  test_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") 
)

print(conf_matrix)
autoplot(conf_matrix, type = "heatmap")
```

### 8. Feature Importance (Interpretability)

This is the key step for the Lasso model. We will extract the coefficients from our final model to see which features it "kept" (gave a non-zero coefficient) and which it "threw away" (set to zero).

```{r vip}
# 12. EXTRACT FEATURE IMPORTANCE
# Extract the fitted model object
fitted_lasso_model <- pull_workflow_fit(final_lasso_fit)

best_lasso_fit <- fit_best(lasso_tune_results)
wflow_fit <- fit(final_lasso_workflow, train_data)

caret::varImp(object = final_lasso_fit)

# Use the 'vip' package to create a variable importance plot
# This shows the non-zero coefficients
print("Plotting Variable Importance...")
vip(best_lasso_fit, num_features = 30) +
  labs(title = "Lasso Variable Importance",
       subtitle = "Features with non-zero coefficients")

# We can also 'tidy' the model to see the coefficients as a data frame
print("Model Coefficients:")
tidy(best_lasso_fit) %>%
  dplyr::filter(estimate != 0) %>%  # Filter to show only non-zero features
  arrange(desc(abs(estimate))) # Sort by importance
```

### 9. Save Model Artifacts

Finally, we save the trained workflow and the performance metrics for our final comparison.

```{r save-artifacts}
# 13. SAVE ARTIFACTS
# Save the final, fitted workflow
saveRDS(
  final_lasso_fit,
  file = here("_models", "final_lasso_fit.rds")
)

# Save the test metrics
saveRDS(
  test_metrics,
  file = here("_models", "lasso_test_metrics.rds")
)

print("Lasso model and metrics saved successfully.")
```
