---
title: "02_lasso_regression.md"
author: "You!"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)

base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',  #log files go here
                      message=FALSE, warning=FALSE,
                      cache = TRUE) 
```  

## OUTLINE OF ANALYSIS 

Step 3 (Implement Lasso): Implement Lasso Logistic Regression. 
  - Use cross-validation to select the optimal regularization parameter (lambda). 
  - The resulting model's coefficients will be analyzed to produce a ranked list of the most important diagnostic features.

```{r}
# Load libraries
library(dplyr)
library(tidymodels)
library(yardstick)
library(ggplot2)
```

## Load Data
```{r}
#load data
raw_dat <- read.csv("../_data/Breast Cancer Wisconsin Data.csv")
str(raw_dat)
```

```{r}
data.m <- raw_dat %>% mutate(
  diag_text = ifelse(diagnosis=="B", "Benign", "Malignant"),
  diag_text = as.factor(diag_text)
) %>%
  select(-id, -diagnosis)

glimpse(data.m)
```

## Pre process numeric vars for LAsso Regression
- If needed


```{r}
numeric_vars <- data.m %>%
  select(where(is.numeric))
```

```{r}
# Standardize numeric variables
numeric_vars_scaled <- as.data.frame(scale(numeric_vars))
```

## Import Lasso Regression Libraries

```{r}
#lasso libraries
library(tidymodels)
#install.packages("glmnet")
library(glmnet)
#install.packages("caret")
library(caret)
```

## Setup Lasso Regression Model
```{r}
#setup lasso regression model
lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")
```

## tidymodels workflow set
```{r}
lasso_workflow <- workflow() %>%
  add_model(lasso_spec) %>%
  add_formula(diag_text ~ .)
```

```{r}
# Cross-validation
set.seed(123)
cv_folds <- vfold_cv(data.m, v = 10, strata = diag_text)
```

```{r}
# Hyperparameter tuning
lasso_grid <- grid_regular(penalty(), levels = 50)

```

```{r}
#show possible evaluation metrics --> YARDSTICK
library(yardstick)
metrics.sel = metric_set(roc_auc, accuracy)
```


```{r}
lasso_tune <- tune_grid(
  lasso_workflow,
  resamples = cv_folds,
  grid = lasso_grid,
  metrics = metrics.sel
)
```

## Create a Confusion Matric for the model predictions
```{r}
#confusion matrix

lasso_best <- lasso_tune %>%
  select_best()
```



