---
title: "02_lasso_regression.md"
author: "You!"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)

base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',  #log files go here
                      message=FALSE, warning=FALSE,
                      cache = TRUE) 
```  

## OUTLINE OF ANALYSIS 

Step # (Implement SVM): Implement Support Vector Machines 
  - Use k-fold cross-validation.
  - Evaluate using metrics/comparisons for SVM performance on BCWDD (Breast Cancer Diagnosis Dataset).
  - Evaluate Model Output, consideration. Indications of critical variables etc...

```{r}
# Load libraries
library(dplyr)
library(tidymodels)
library(recipes)
library(yardstick)
library(ggplot2)
```

## Load Data
```{r}
#load data
raw_dat <- read.csv("../_data/Breast Cancer Wisconsin Data.csv")
str(raw_dat)
```

```{r}
data.m <- raw_dat %>% mutate(
  diag_text = ifelse(diagnosis=="B", "Benign", "Malignant"),
  diag_text = as.factor(diag_text)
) %>%
  select(-id, -diagnosis)

glimpse(data.m)
```

## Pre process Variables as Required


## Import any Specific SVM packages needed
- Check against 00_setup_libraries.rmd for installs. 
```{r}

```

## Setup SVM Engine/Model



## Evaluation Metrics
```{r}
#show possible evaluation metrics --> YARDSTICK
library(yardstick)
#set metrics for SVM
metrics.sel = metric_set(roc_auc, accuracy)
```


## Create a Confusion Matric for the model predictions



