---
title: "02_lasso_regression.md"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)
```  

This script implements Step 4 of the project proposal: fitting and tuning a non-linear Support Vector Machine (SVM) model with a Radial Basis Function (RBF) kernel.

```{r libraries}

library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(kernlab)     # The 'engine' for SVM calculations
library(conflicted)  # To manage function name conflicts
library(vip)         #variables of importance

# Set a seed for reproducibility
set.seed(8888) 
```

## OUTLINE OF ANALYSIS 

Model 3 (SVM): Implement Support Vector Machines 
  - Use k-fold cross-validation.
  - Evaluate using metrics/comparisons for SVM performance on BCWDD (Breast Cancer Diagnosis Dataset).
  - Evaluate Model Output, consideration. Indications of critical variables etc...

### 1. Load Data

```{r}
#load data
cancer_data <- readRDS(here("_data", "cancer_data_processed.rds")) #processed in00 data_wrangling script.
str(cancer_data)
```
### 2. Data Splitting

We use the same stratified split parameters and seed as the SVM script to ensure the train/test sets are identical.

```{r data-split}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

### 3. Preprocessing Recipe

We define our preprocessing steps. For an SVM, it is **critical** that we `step_scale()` and `step_center()` all numeric predictors. The recipe will be *trained* on the training data and *applied* to both train and test sets.

```{r recipe}
# 4. PREPROCESSING RECIPE
# We are predicting 'diagnosis' using all other variables (.).
# The only predictor we *don't* want is 'id', so we remove it.
# We then center and scale ALL numeric predictors.
svm_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it as a predictor
  step_normalize(all_numeric_predictors()) # Centers and scales

# You can check the recipe
# summary(svm_recipe)
```

### 4. Cross-Validation Folds

We create 10-fold cross-validation folds from our *training data*. This will be used to tune the model's hyperparameters.

```{r cv-folds}
# 5. CROSS-VALIDATION FOLDS
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### Working out the Model fit code

```{r}
#run a single SVM fit
svm_fit <- svm_rbf(cost = 1, rbf_sigma = 0.1) %>%
  set_mode("classification") %>%
  set_engine("kernlab") %>%
  #fit(diagnosisdata = train_data)
  fit(diagnosis ~ ., data = train_data %>% select(-id))

summary(svm_fit)
preds <- predict(object = svm_fit, train_data)
compare <- data.frame(truth = train_data$diagnosis, estimate = preds$.pred_class)
yardstick::accuracy(data = compare, truth = truth, estimate = estimate)
```

### 5. Model Specification & Tuning
Here we define the SVM model, specify which hyperparameters we want to tune (`cost` and `rbf_sigma`), and create a tuning grid.

```{r model-spec}
# 6. MODEL SPECIFICATION
# We specify a radial basis function (RBF) kernel SVM
# We set the 'engine' to 'kernlab'
# We will tune 'cost' (the C parameter) and 'rbf_sigma' (gamma)
svm_spec <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

# 7. CREATE WORKFLOW
# Combine the recipe and model spec into a single workflow object
svm_workflow <- workflow() %>%
  add_recipe(svm_recipe) %>%
  add_model(svm_spec)

# 8. HYPERPARAMETER TUNING
# Create a grid of hyperparameters to try
# We will use a space-filling design to get 25 combinations
svm_grid <- grid_latin_hypercube(
  cost(),
  rbf_sigma(),
  size = 25
)

ctrl <- control_grid(save_workflow = TRUE)

# Run the tuning process across all CV folds
# This will try all 25 hyperparameter combinations on our 10 folds
svm_tune_results <- tune_grid(
  svm_workflow,
  resamples = cv_folds,
  grid = svm_grid,
  control=ctrl)

# Show the top 5 best-performing hyperparameter sets
show_best(svm_tune_results, metric = "accuracy")

# Select the single best set of parameters based on accuracy
best_svm_params <- select_best(svm_tune_results, metric = "accuracy")
```

### 6. Finalize and Evaluate Model

Now we update our workflow with the *best* parameters and fit it one last time to the *full training set*. Then, we evaluate its performance on the *held-out test set*.

```{r final-fit}
# 9. FINALIZE WORKFLOW
# Update the workflow with the best parameters we just found
final_svm_workflow <- finalize_workflow(
  svm_workflow,
  best_svm_params
)

# 10. FIT ON TRAIN, EVALUATE ON TEST
# Use the 'last_fit' function to:
# 1. Fit the final workflow on the *entire training set*
# 2. Evaluate that fitted model on the *testing set*
final_svm_fit <- last_fit(
  final_svm_workflow,
  data_split
)
```

### 7. Review Performance Metrics
Compare *tuned* model to test data
```{r metrics}
# Get the MODEL METRICS metrics from our test-set evaluation
test_metrics <- collect_metrics(final_svm_fit)
print(test_metrics)
```

## Accuracy, Sensitivity, Recall Metrics
```{r}
# Get the PERFORMANCE metrics predictions to build a confusion matrix
test_predictions <- collect_predictions(final_svm_fit)
# Metrics from ML Course
ml_mets <- metric_set(accuracy, precision, recall)
ml_mets(test_predictions, truth = diagnosis, estimate = .pred_class)
```

## Generate and print the confusion matrix
```{r}

# Note: In a confusion matrix, 'M' (Malignant) is our "positive" class.
# We set 'M' to be the first level to ensure this.
conf_matrix <- conf_mat(
  test_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") # Explicitly set positive class
)

print(conf_matrix)
```

```{r}
#Heatmap
autoplot(conf_matrix, type = "heatmap")
```

### 8. Save Model Artifacts

```{r save-artifacts}
# 12. SAVE ARTIFACTS
# Save the final, fitted workflow
saveRDS(
  list("model" = final_svm_fit, "metrics" = test_metrics, "preds" = test_predictions),
  file = here("_models", "final_svm_output.rds")
)
```


