---
title: "02_lasso_regression.md"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)
```  

This script implements Step 4 of the project proposal: fitting and tuning a non-linear Support Vector Machine (SVM) model with a Radial Basis Function (RBF) kernel.

```{r libraries}
library(dplyr)
library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(kernlab)     # The 'engine' for SVM calculations
library(conflicted)  # To manage function name conflicts
library(vip)         #variables of importance

# Set a seed for reproducibility
set.seed(8888) 
```


**OUTLINE OF ANALYSIS **

Model 3 (SVM): Implement Support Vector Machines 
  - Use k-fold cross-validation.
  - Evaluate using metrics/comparisons for SVM performance on BCWDD (Breast Cancer Diagnosis Dataset).
  - Evaluate Model Output, consideration. Indications of critical variables etc...

## Support Vector Machine (SVM) Implementation
### Methodological Note: 
  SVM searches for an optimal hyperplane that separates the two classes (Benign vs. Malignant) with the maximum margin. Since the tumor data is not linearly separable (complex overlaps between classes), we use the Kernel Trick. 
  Optimization Problem: We minimize the following cost function:
  $$ \min_{w, b, \xi} \left( \frac{1}{2} ||w||^2 + C \sum_{i=1}^{N} \xi_i \right) $$
  Subject to:$$ y_i (w \cdot \phi(x_i) + b) \geq 1 - \xi_i $$
  Where:$\phi(x_i)$ maps the input vectors into a higher-dimensional space.$\xi_i$ are slack variables allowing for some misclassification (controlled by $C$).
  The RBF Kernel handles the mapping implicitly: 
  $K(x_i, x_j) = \exp(-\gamma ||x_i - x_j||^2)$.
  
  
### 1. Load Data

```{r}
#load data
cancer_data <- readRDS(here("_data", "cancer_data_processed.rds")) #processed in00 data_wrangling script.
str(cancer_data)
```
### 2. Data Splitting

We use the same stratified split parameters and seed as the SVM script to ensure the train/test sets are identical.

```{r data-split}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

### 3. Preprocessing Recipe

We define our preprocessing steps. For an SVM, it is **critical** that we `step_scale()` and `step_center()` all numeric predictors. The recipe will be *trained* on the training data and *applied* to both train and test sets.

Critical Step: SVMs rely on distance calculations (Euclidean distance in the kernel). If features are not scaled (e.g., area_mean ~ 1000 vs smoothness ~ 0.1), the kernel will be dominated by the large variables. Normalization is mandatory.

```{r recipe}
# 4. PREPROCESSING RECIPE
# We are predicting 'diagnosis' using all other variables (.).
# The only predictor we *don't* want is 'id', so we remove it.
# We then center and scale ALL numeric predictors.
svm_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it as a predictor
  step_normalize(all_numeric_predictors()) # Centers and scales

# You can check the recipe
# summary(svm_recipe)
```

### 4. Cross-Validation Folds

We create 10-fold cross-validation folds from our *training data*. This will be used to tune the model's hyperparameters.

```{r cv-folds}
# 5. CROSS-VALIDATION FOLDS
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### Working out the Model fit code

```{r}
#run a single SVM fit
svm_fit <- svm_rbf(cost = 1, rbf_sigma = 0.1) %>%
  set_mode("classification") %>%
  set_engine("kernlab") %>%
  #fit(diagnosisdata = train_data)
  fit(diagnosis ~ ., data = train_data %>% select(-id))

summary(svm_fit)
preds <- predict(object = svm_fit, train_data)
compare <- data.frame(truth = train_data$diagnosis, estimate = preds$.pred_class)
yardstick::accuracy(data = compare, truth = truth, estimate = estimate)
```

### 5. Model Specification & Tuning
Here we define the SVM model, specify which hyperparameters we want to tune (`cost` and `rbf_sigma`), and create a tuning grid.

Hyperparameters:
  Cost (cost): Controls the penalty for misclassification. High cost = strict margin (risk of overfitting). Low cost = soft margin.
  RBF Sigma (rbf_sigma): Controls the width of the Gaussian kernel. High sigma = complex, wiggly decision boundary.

```{r model-spec}
# 6. MODEL SPECIFICATION
# We specify a radial basis function (RBF) kernel SVM
# We set the 'engine' to 'kernlab'
# We will tune 'cost' (the C parameter) and 'rbf_sigma' (gamma)
svm_spec <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

# 7. CREATE WORKFLOW
# Combine the recipe and model spec into a single workflow object
svm_workflow <- workflow() %>%
  add_recipe(svm_recipe) %>%
  add_model(svm_spec)

# 8. HYPERPARAMETER TUNING
# Create a grid of hyperparameters to try
# We will use a space-filling design to get 25 combinations
svm_grid <- grid_latin_hypercube(
  cost(),
  rbf_sigma(),
  size = 25
)

ctrl <- control_grid(save_workflow = TRUE)

# Run the tuning process across all CV folds
# This will try all 25 hyperparameter combinations on our 10 folds
svm_tune_results <- tune_grid(
  svm_workflow,
  resamples = cv_folds,
  grid = svm_grid,
  control=ctrl)

# Show the top 5 best-performing hyperparameter sets
show_best(svm_tune_results, metric = "accuracy")
```

```{r}
# Select the single best set of parameters based on accuracy
best_svm_params <- select_best(svm_tune_results, metric = "accuracy")
best_svm_params
```

### 6. Finalize and Evaluate Model

Now we update our workflow with the *best* parameters and fit it one last time to the *full training set*. Then, we evaluate its performance on the *held-out test set*.

```{r final-fit}
# 9. FINALIZE WORKFLOW
# Update the workflow with the best parameters we just found
final_svm_workflow <- finalize_workflow(
  svm_workflow,
  best_svm_params
)

# 10. FIT ON TRAIN, EVALUATE ON TEST
# Use the 'last_fit' function to:
# 1. Fit the final workflow on the *entire training set*
# 2. Evaluate that fitted model on the *testing set*
final_svm_fit <- last_fit(
  final_svm_workflow,
  data_split
)
```

### 7. Review Performance Metrics
Compare *tuned* model to test data
```{r metrics}
# Get the MODEL METRICS metrics from our test-set evaluation
svm_test_metrics <- collect_metrics(final_svm_fit)
print(svm_test_metrics)
```

## Accuracy, Sensitivity, Recall Metrics
```{r}
# Get the PERFORMANCE metrics predictions to build a confusion matrix
svm_predictions <- collect_predictions(final_svm_fit)
# Metrics from ML Course
ml_mets <- metric_set(accuracy, precision, recall, f_meas)
ml_mets(svm_predictions, truth = diagnosis, estimate = .pred_class)
```

## Generate and print the confusion matrix
```{r}

# Note: In a confusion matrix, 'M' (Malignant) is our "positive" class.
# We set 'M' to be the first level to ensure this.
conf_matrix <- conf_mat(
  svm_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") # Explicitly set positive class
)

print(conf_matrix)
```

```{r}
#Heatmap
autoplot(conf_matrix, type = "heatmap")
```

# plot ROC-AUC Curves
```{r}
svm_probs <- final_svm_workflow %>% 
  fit(data = train_data) %>%
  predict(new_data = test_data, type = "prob") %>%
  bind_cols(test_data %>% select(diagnosis))

svm_probs %>% 
  roc_curve(diagnosis, .pred_B) %>%
  autoplot() +
  labs(
    title = "ROC Curve: SVM (RBF Kernel)",
    subtitle = "Performance of the Non-Linear Support Vector Machine"
  )


# #plot roc-auc for the SVM Model
# svm_predictions %>%
#   #pr_curve(diagnosis, .pred_B) %>%
#   roc_curve(diagnosis, .pred_M) %>%
#   autoplot() +
#   labs(
#     title = "ROC Curve: SVM (RBF Kernel)",
#     subtitle = "Performance of the Non-Linear Support Vector Machine"
#   )

```



```{r}
# Precision-Recall Curve
# The RBF kernel often produces excellent separation, potentially yielding
# higher Area Under the Curve (AUC) than linear methods.
svm_predictions %>%
  pr_curve(diagnosis, .pred_M) %>%
  autoplot() +
  labs(
    title = "Precision-Recall Curve: SVM (RBF Kernel)",
    subtitle = "Performance of the Non-Linear Support Vector Machine"
  )
```


### 8. Save Model Artifacts

```{r save-artifacts}
# 12. SAVE ARTIFACTS
# Save the final, fitted workflow
saveRDS(
  list("model" = final_svm_fit, "metrics" = svm_test_metrics, "preds" = svm_predictions),
  file = here("_models", "final_svm_output.rds")
)
```


