---
title: "02_lasso_regression.md"
author: "You!"
date: "2025-10-31"
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
source("../_templates/utility_functions.R")
.default_slick_knit_ops(verbose=TRUE)
# 
# base_dir <- "../" # where output should default to
# base_url <- "" # keep as is
# fig_path <- "_plot_images/" # where image output should go
# 
# knitr::opts_knit$set(base.dir = base_dir)
# knitr::opts_chunk$set(fig.path = fig_path,
#                       cache.path = '../cache/',  #log files go here
#                       message=FALSE, warning=FALSE,
#                       cache = TRUE) 
```  

This script implements Step 4 of the project proposal: fitting and tuning a non-linear Support Vector Machine (SVM) model with a Radial Basis Function (RBF) kernel.

```{r libraries}

library(tidymodels)  # For the core modeling workflow
library(tidyverse)   # For general data manipulation and loading
library(here)        # For robust file paths
library(kernlab)     # The 'engine' for SVM calculations
library(conflicted)  # To manage function name conflicts

# Set a seed for reproducibility
set.seed(8888) 
```

## OUTLINE OF ANALYSIS 

Model 3 (SVM): Implement Support Vector Machines 
  - Use k-fold cross-validation.
  - Evaluate using metrics/comparisons for SVM performance on BCWDD (Breast Cancer Diagnosis Dataset).
  - Evaluate Model Output, consideration. Indications of critical variables etc...

### 1. Load and Prepare Data

Load the cleaned data from the `01_data_wrangling.Rmd` script. The most important step here is converting our `diagnosis` variable to a **factor**, which `tidymodels` requires for classification.

```{r load-data}
# 2. LOAD DATA
# Load the cleaned data from the _data folder
raw_BCWDD <- read_csv(here("_data", "Breast Cancer Wisconsin Data.csv"))

names(raw_BCWDD) <- gsub(names(raw_BCWDD), pattern = " ", replacement = "_")

cancer_data <- raw_BCWDD %>% 
  mutate(
    # Convert the diagnosis column to a factor
    # This is ESSENTIAL for classification models in tidymodels
    diagnosis = as_factor(diagnosis)) %>% 
    #diagnosis.fac_lab = factor(diagnosis, levels = c("B", "M"), labels=c("B", "M"))) # 'B' = Benign, 'M' = Malignant
  select(-33) #remove unnamed column if present


#str(cancer_data$diagnosis)
#str(cancer_data$diagnosis.fac_lab)

glimpse(cancer_data)
```

### 2. Data Splitting

We will perform a stratified split to ensure our training and testing sets both have a similar proportion of Malignant ('M') and Benign ('B') cases.

```{r data-split}
# 3. DATA SPLITTING
# Create the stratified data split (75% training, 25% testing)
data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)

# Create the training and testing data frames
train_data <- training(data_split)
test_data  <- testing(data_split)

# Check the proportions in each set
train_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
test_data %>% count(diagnosis) %>% mutate(prop = n/sum(n))
```

### 3. Preprocessing Recipe

We define our preprocessing steps. For an SVM, it is **critical** that we `step_scale()` and `step_center()` all numeric predictors. The recipe will be *trained* on the training data and *applied* to both train and test sets.

```{r recipe}
# 4. PREPROCESSING RECIPE
# We are predicting 'diagnosis' using all other variables (.).
# The only predictor we *don't* want is 'id', so we remove it.
# We then center and scale ALL numeric predictors.
svm_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  update_role(id, new_role = "ID") %>% # Keep ID but don't use it as a predictor
  step_normalize(all_numeric_predictors()) # Centers and scales

# You can check the recipe
# summary(svm_recipe)
```

### 4. Cross-Validation Folds

We create 10-fold cross-validation folds from our *training data*. This will be used to tune the model's hyperparameters.

```{r cv-folds}
# 5. CROSS-VALIDATION FOLDS
# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)
```

### Working out the Model fit code

```{r}
#run a single SVM fit
svm_fit <- svm_rbf(cost = 1, rbf_sigma = 0.1) %>%
  set_mode("classification") %>%
  set_engine("kernlab") %>%
  #fit(diagnosisdata = train_data)
  fit(diagnosis ~ ., data = train_data %>% select(-id))

summary(svm_fit)
preds <- predict(object = svm_fit, train_data)
compare <- data.frame(truth = train_data$diagnosis, estimate = preds$.pred_class)
yardstick::accuracy(data = compare, truth = truth, estimate = estimate)
```

### 5. Model Specification & Tuning
Here we define the SVM model, specify which hyperparameters we want to tune (`cost` and `rbf_sigma`), and create a tuning grid.

```{r model-spec}
# 6. MODEL SPECIFICATION
# We specify a radial basis function (RBF) kernel SVM
# We set the 'engine' to 'kernlab'
# We will tune 'cost' (the C parameter) and 'rbf_sigma' (gamma)
svm_spec <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

# 7. CREATE WORKFLOW
# Combine the recipe and model spec into a single workflow object
svm_workflow <- workflow() %>%
  add_recipe(svm_recipe) %>%
  add_model(svm_spec)

# 8. HYPERPARAMETER TUNING
# Create a grid of hyperparameters to try
# We will use a space-filling design to get 25 combinations
svm_grid <- grid_latin_hypercube(
  cost(),
  rbf_sigma(),
  size = 25
)

ctrl <- control_grid(save_workflow = TRUE)

# Run the tuning process across all CV folds
# This will try all 25 hyperparameter combinations on our 10 folds
svm_tune_results <- tune_grid(
  svm_workflow,
  resamples = cv_folds,
  grid = svm_grid,
  control=ctrl)

# Show the top 5 best-performing hyperparameter sets
show_best(svm_tune_results, metric = "accuracy")

# Select the single best set of parameters based on accuracy
best_svm_params <- select_best(svm_tune_results, metric = "accuracy")
```

### 6. Finalize and Evaluate Model

Now we update our workflow with the *best* parameters and fit it one last time to the *full training set*. Then, we evaluate its performance on the *held-out test set*.

```{r final-fit}
# 9. FINALIZE WORKFLOW
# Update the workflow with the best parameters we just found
final_svm_workflow <- finalize_workflow(
  svm_workflow,
  best_svm_params
)

# 10. FIT ON TRAIN, EVALUATE ON TEST
# Use the 'last_fit' function to:
# 1. Fit the final workflow on the *entire training set*
# 2. Evaluate that fitted model on the *testing set*
final_svm_fit <- last_fit(
  final_svm_workflow,
  data_split
)
```

### 7. Review Performance Metrics

Let's see how our final model performed on the test data. We will look at the key metrics requested in the proposal.
- Predict on the test set. 
- Evaluate metrics against preds and actual diagnoses.

```{r metrics}
best_svm_fit <- fit_best(svm_tune_results) #best fit model
best_predictions <- predict(best_svm_fit, test_data) %>% 
  bind_cols(test_data %>% select(diagnosis))

# Performance Metrics - How Does the Model actually do on the test set.
# Metrics from ML Course
ml_mets <- metric_set(accuracy, precision, recall)

svm_ml_metrics <- ml_mets(best_predictions,
        truth = diagnosis, estimate = .pred_class, metrics = ml_mets) %>% pivot_wider(names_from = .metric, values_from = .estimate)

svm_ml_metrics

f1_score <- function(metrics_data){
  2 * metrics_data$precision * metrics_data$recall / (metrics_data$precision + metrics_data$recall)
  }

f1_score(svm_ml_metrics)

```

```{r}
# 11. COLLECT METRICS
# Get the metrics from our test-set evaluation
test_metrics <- collect_metrics(final_svm_fit)
print(test_metrics)

# Get the predictions to build a confusion matrix
test_predictions <- collect_predictions(final_svm_fit)

#metrics from preds
# Metrics from ML Course
ml_mets <- metric_set(accuracy, precision, recall)

#svm_metrics <- test_predictions %>% mutate(diagnosis = factor(diagnosis, levels = c("B", "M"))) %>%
#  metrics(truth = diagnosis, estimate = .pred_class)

# Generate and print the confusion matrix
# Note: In a confusion matrix, 'M' (Malignant) is our "positive" class.
# We set 'M' to be the first level to ensure this.
conf_matrix <- conf_mat(
  test_predictions,
  truth = diagnosis,
  estimate = .pred_class,
  options = list(positive = "M") # Explicitly set positive class
)

print(conf_matrix)

# You can also autoplot the confusion matrix
autoplot(conf_matrix, type = "heatmap")
```

### 8. Save Model Artifacts

Finally, we save the trained workflow (which includes the preprocessing recipe) and the performance metrics. This allows us to load them later in our final comparison script without re-running this entire file.

```{r save-artifacts}
# 12. SAVE ARTIFACTS
# Save the final, fitted workflow
saveRDS(
  final_svm_fit,
  file = here("_models", "final_svm_fit.rds")
)

# Save the test metrics
saveRDS(
  test_metrics,
  file = here("_models", "svm_test_metrics.rds")
)

print("SVM model and metrics saved successfully.")
```


