---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r setup, include=FALSE}
#title: "04_Compare_Models.md"
source("../_templates/utility_functions.R")
.default_slick_knit_ops()
library(kableExtra)
library(knitr)
library(dplyr)         #For data manipulation
```  

## Workflow Reproducibility and Model Comparison
This script outlines the 'tidymodels' workflow used for the project. 

```{r libraries}
#library(tidymodels)   For the core modeling framework
#library(tidyverse)    For data manipulation and visualization
#library(vip)          Used for variable importance plots
#library(kernlab)      SVM Library
```
### (A) Dataset Description
Breast Cancer Wisconsin (Diagnostic) Dataset (BCWDD)
 - Dimensions: 568 observations (rows) and 32 variables (columns).
 - Target Variable: 'diagnosis' (Factor w/ 2 levels: "B" for Benign, "M" for Malignant).
 - Predictors: 30 numerical features (e.g., radius_mean, texture_mean, area_worst).
 - ID Variable: 'id' (removed during preprocessing).

### (B) Tidymodels Approach for Reproducibility

```{r tidymodels_workflow, eval=FALSE}
#Set a single seed for all processes to ensure reproducibility
 set.seed(8888)

# 1. Data Splitting
#  The data was split into training (75%) and testing (25%) sets.
#  Stratification on 'diagnosis' was used to maintain class balance.
 data_split <- initial_split(cancer_data, prop = 0.75, strata = diagnosis)
 train_data <- training(data_split)
 test_data  <- testing(data_split)

# 2. Cross-Validation Folds
#   10-fold cross-validation was created from the *training data*
#   This same set of folds ('cv_folds') was used to tune all three models.
  cv_folds <- vfold_cv(train_data, v = 10, strata = diagnosis)

# 3. Preprocessing Recipe
#   A single, consistent recipe was defined to be applied to all models.
#   This ensures all models received the exact same processed data.
  norm_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
   step_rm(id) %>%   The 'id' column is not a predictor
   step_normalize(all_numeric_predictors())  Center and scale all 30 predictors


# 4. Model Specifications for Tuning
# These are the "blueprints" for the models, with parameters set to 'tune()'.
 
  # --- Model 1: Random Forest ---
 rf_spec_tune <- rand_forest(
   mtry = tune(),   Number of predictors to sample
   trees = 1000,    1000 trees is generally sufficient
   min_n = tune()   Minimum node size
 ) %>%
   set_engine("ranger", importance = "permutation") %>%
   set_mode("classification")

  # --- Model 2: Lasso (L1) Logistic Regression ---
 lasso_spec_tune <- logistic_reg(
   penalty = tune(),  The L1 regularization penalty (lambda)
   mixture = 1        mixture = 1 specifies a pure Lasso model
 ) %>%
   set_engine("glmnet") %>%
   set_mode("classification")

  # --- Model 3: Support Vector Machine (RBF Kernel) ---
 svm_spec_tune <- svm_rbf(
   cost = tune(),       The cost of misclassification
   rbf_sigma = tune()   The gamma/sigma of the RBF kernel
 ) %>%
   set_engine("kernlab") %>%
   set_mode("classification")

# 5. Model Fitting Routine (Example)
# Each model was then tuned using a workflow, tune_grid, and the 'cv_folds'.
  # The example below is for the Random Forest.

  rf_workflow_tune <- workflow() %>%
    add_model(rf_spec_tune) %>%
    add_recipe(norm_recipe)
 
  rf_tune_results <- tune_grid(
    rf_workflow_tune,
    resamples = cv_folds,
    grid = 20  Test 20 candidate hyperparameter combinations
  )
 
  #  The 'best' model was selected, finalized, and fit one last time
  #  on the full training set and evaluated on the test set.
  best_rf_params <- select_best(rf_tune_results, metric = "accuracy")
 
  final_rf_fit <- last_fit(
    finalize_workflow(rf_workflow_tune, best_rf_params),
    data_split
  )
```

### (C) Final Tables Comparing Model Outcomes
The data below is hard-coded from the final, verified results
from running the 'last_fit()' evaluation on the test set (n=142).

```{r echo=FALSE}
# --- Table 1: Final Performance Metrics ---
#message("--- Final Performance Metrics (Test Set) ---")
comparison_metrics <- tibble(
  model = c("Random Forest", "Lasso Regression", "SVM (RBF)"),
  accuracy = c(0.958, 0.951, 0.958),
  precision = c(0.980, 0.979, 0.980),
  recall = c(0.906, 0.887, 0.906)
)
kable(comparison_metrics) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "center") %>%
  kableExtra::add_header_above(c("Final Performance Metrics (Test Set)"=4))
```

```{r echo=FALSE}
#### Table 2: Final Confusion Matrix Comparison
#message("\n--- Final Confusion Matrix Summary (Test Set) ---")
# Rows represent 'Actual' diagnosis, Columns 'Predicted' diagnosis.
confusion_summary <- tibble(
  model = c("Random Forest", "Lasso Regression", "SVM (RBF)"),
  true_negative_B = c(88, 88, 88),   # Predicted Benign, Actual Benign
  false_positive_M = c(1, 1, 1),    # Predicted Malignant, Actual Benign
  false_negative_B = c(5, 6, 5),    # Predicted Benign, Actual Malignant
  true_positive_M = c(48, 47, 48)   # Predicted Malignant, Actual Malignant
)
kable(confusion_summary) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "center") %>%
  kableExtra::add_header_above(c("Final Confusion Matrix (Test Set)"=5))
```

```{r echo=FALSE}
#### Table 3: Best Tuned Hyperparameters
#message("\n--- Best Tuned Hyperparameters (from 10-Fold CV) ---")
best_params <- tibble(
  model = c("Random Forest", "Lasso Regression", "SVM (RBF)"),
  parameter_1 = c("mtry = 10", "penalty = 0.00356", "cost = 21.5"),
  parameter_2 = c("min_n = 6", NA, "rbf_sigma = 0.000318")
)
kable(best_params) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "center") %>%
  kableExtra::add_header_above(c("Final Confusion Matrix (Test Set)"=3))
```
