---
title: "ML_Optimization_Problem2"
date: "2025-12-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(combinat)
library(mvtnorm)
```

# STSCI 6520 Homework 2 - Problem 2
K.CHAN

## Gaussian Mixture Models, K-Means, and EM Algorithm
- The Expectation-Maximization (EM) algorithm is an iterative method to find maximum likelihood estimates when data is missing or hidden. 
- In a Gaussian Mixture Model (GMM), the "hidden" data is the knowledge of which Gaussian component generated which data point.
- The Algorithm Steps:Initialization: Start with initial guesses for weights ($w$), means ($\mu$), and covariances ($\Sigma$). 
- In this problem, we use the output of K-means for this.E-Step (Expectation): Calculate the "responsibility" $\gamma_{ik}$, which is the probability that point $i$ belongs to cluster $k$, given the current parameters.
$$\gamma_{ik} = \frac{w_k \cdot \mathcal{N}(x_i | \mu_k, \Sigma_k)}{\sum_{j=1}^{3} w_j \cdot \mathcal{N}(x_i | \mu_j, \Sigma_j)}$$
- M-Step (Maximization): Update the parameters to maximize the likelihood using the responsibilities from the E-step.
- New Weights: $w_k^{new} = \frac{1}{n} \sum_{i=1}^n \gamma_{ik}$
- New Means: $\mu_k^{new} = \frac{\sum_{i=1}^n \gamma_{ik} x_i}{\sum_{i=1}^n \gamma_{ik}}$
- New Covariances: $\Sigma_k^{new} = \frac{\sum_{i=1}^n \gamma_{ik} (x_i - \mu_k^{new})(x_i - \mu_k^{new})^T}{\sum_{i=1}^n \gamma_{ik}}$
Repeat until convergence.


```{r}
set.seed(8888)
#1. data gen
# Parameters
true_weights <- c(0.6, 0.2, 0.2)
mu <- list(
  c(0, 0),
  c(-1.6, 1.4),
  c(1.6, 1.4)
)
sigma <- list(
  diag(1, 2),
  diag(0.25, 2),
  diag(0.25, 2)
)

n <- 200

# Function to generate data
generate_gmm_data <- function(n, weights, means, sigmas) {
  # Sample cluster assignments based on weights
  z <- sample(1:3, n, replace = TRUE, prob = weights)
  
  X <- matrix(0, nrow = n, ncol = 2)
  for (i in 1:n) {
    k <- z[i]
    X[i, ] <- rmvnorm(1, mean = means[[k]], sigma = sigmas[[k]])
  }
  
  return(list(X = X, labels = z))
}

data <- generate_gmm_data(n, true_weights, mu, sigma)
X <- data$X
true_labels <- data$labels

# Plot Setup
plot_clusters <- function(X, labels, title) {
  colors <- c("red", "blue", "green")
  plot(X, col = colors[labels], pch = 19, main = title, 
       xlab = "X1", ylab = "X2", asp = 1)
  grid()
}
#plot_clusters(X, true_labels, "True Cluster Assignments")
```

```{r}
# Since K-means/EM are unsupervised, Cluster 1 might assume the ID "2".
# We must permute predicted labels to match ground truth to minimize error.
get_best_match_error <- function(true_labs, pred_labs) {
  library(combinat) # For permn, or we hardcode the 6 permutations for 3 clusters
  
  perms <- list(
    c(1, 2, 3), c(1, 3, 2),
    c(2, 1, 3), c(2, 3, 1),
    c(3, 1, 2), c(3, 2, 1)
  )
  
  best_error <- 1.0
  best_mapping <- NULL
  
  for (p in perms) {
    # Map prediction k -> p[k]
    mapped_preds <- p[pred_labs]
    error <- mean(mapped_preds != true_labs)
    
    if (error < best_error) {
      best_error <- error
      best_mapping <- mapped_preds
    }
  }
  
  return(list(error = best_error, mapped_labels = best_mapping))
}
```

## 2. K-Means Implementation

```{r}
run_kmeans <- function(X, K = 3, max_iter = 100) {
  n <- nrow(X)
  centroids <- X[sample(1:n, K), ] #sample RANDOMIZES the points.
  assignments <- rep(0, n)
  
  for (iter in 1:max_iter) {
    # Assignment Step
    dists <- matrix(0, n, K)
    for (k in 1:K) {
      # Squared Euclidean distance
      dists[, k] <- rowSums((sweep(X, 2, centroids[k, ])) ^ 2)
    }
    new_assignments <- apply(dists, 1, which.min)
    
    # Check convergence
    if (all(assignments == new_assignments)) break
    assignments <- new_assignments
    
    # Update Step
    for (k in 1:K) {
      if (sum(assignments == k) > 0) {
        centroids[k, ] <- colMeans(X[assignments == k, , drop = FALSE])
      }
    }
  }
  return(list(assignments = assignments, centroids = centroids))
}

km_res <- run_kmeans(X)
km_eval <- get_best_match_error(true_labels, km_res$assignments)
```

## 2A and B SOLUTION 
- K-Means Fraction of Mistakes: 0.27 
```{r}
plot_clusters(X, km_eval$mapped_labels, paste0("K-Means Clustering (Error: ", round(km_eval$error, 3), ")"))
```

# 2c. EM Algorithm Implementation
```{r}

# Initialize EM parameters using K-means results
init_em <- function(X, km_assignments, K = 3) {
  w <- numeric(K)
  mu <- list()
  sigma <- list()
  
  for (k in 1:K) {
    mask <- (km_assignments == k)
    w[k] <- mean(mask) # Fraction of points in cluster k
    mu[[k]] <- colMeans(X[mask, , drop = FALSE])
    sigma[[k]] <- cov(X[mask, , drop = FALSE])
    
    # Regularize covariance to avoid singularity if cluster is too small
    if (sum(mask) < 2) sigma[[k]] <- diag(1, 2) 
  }
  return(list(w = w, mu = mu, sigma = sigma))
}

run_em <- function(X, init_params, K = 3, max_iter = 100, tol = 1e-6) {
  n <- nrow(X)
  w <- init_params$w
  mu <- init_params$mu
  sigma <- init_params$sigma
  
  log_lik_hist <- c()
  
  for (iter in 1:max_iter) {
    # --- E-Step ---
    # Calculate responsibilities gamma_ik
    gammas <- matrix(0, n, K)
    for (k in 1:K) {
      gammas[, k] <- w[k] * dmvnorm(X, mean = mu[[k]], sigma = sigma[[k]])
    }
    
    # Normalize rows to sum to 1
    row_sums <- rowSums(gammas)
    log_lik <- sum(log(row_sums)) # Track log likelihood
    gammas <- gammas / row_sums
    
    # Check convergence
    if (iter > 1 && abs(log_lik - tail(log_lik_hist, 1)) < tol) break
    log_lik_hist <- c(log_lik_hist, log_lik)
    
    # --- M-Step ---
    N_k <- colSums(gammas) # Effective number of points in each cluster
    
    for (k in 1:K) {
      # Update weights
      w[k] <- N_k[k] / n
      
      # Update means
      # Weighted sum of X
      mu[[k]] <- colSums(gammas[, k] * X) / N_k[k]
      
      # Update covariances
      # Weighted covariance matrix
      # (X - mu_k)
      diff <- sweep(X, 2, mu[[k]])
      # Manually compute weighted outer product sum
      # Sigma_k = sum(gamma_ik * (x_i - mu_k)(x_i - mu_k)^T) / N_k
      sig_sum <- matrix(0, 2, 2)
      for (i in 1:n) {
        sig_sum <- sig_sum + gammas[i, k] * (diff[i, ] %*% t(diff[i, ]))
      }
      sigma[[k]] <- sig_sum / N_k[k]
    }
  }
  
  # Final hard assignments for error calculation
  final_assignments <- apply(gammas, 1, which.max)
  
  return(list(w = w, mu = mu, sigma = sigma, 
              assignments = final_assignments, gammas = gammas))
}
```

## 2C SOLUTION
```{r}
# Running EM Func above
# Note: Func using the *raw* K-means assignments for initialization, not the mapped ones.
# The algorithm will adapt the parameters to fit the data distribution.
init_params <- init_em(X, km_res$assignments)
em_res <- run_em(X, init_params)

# Evaluate EM Error
em_eval <- get_best_match_error(true_labels, em_res$assignments)
```

EM Algorithm Fraction of Mistakes: 0.13 
Compare Estimated Parameters to True Parameters
--- Parameter Comparison ---
True Weights: 0.6 0.2 0.2 
Est. Weights: 0.22 0.24 0.53 

EM improves performance of k-means algorithm significantly!
```{r}
plot_clusters(X, em_eval$mapped_labels, paste0("EM Clustering (Error: ", round(em_eval$error, 3), ")"))
```
